A deeper look into bug fixes: Patterns, replacements, deletions, and additions

### 2016-02-19 MSRChallenge2016.tex
Revised Sec.4

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Mauricio Soto\\
       \affaddr{School of Computer Science}\\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{mauriciosoto@cmu.edu}
       % 3rd. author
\alignauthor Ferdian Thung\\
       \affaddr{Singapore Management University}\\
       \affaddr{Singapore}\\
       \email{ferdiant.2013@smu.edu.sg}
% 2nd. author
\alignauthor Chu-Pan Wong   \\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{chupanw@cs.cmu.edu}       
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor
Claire Le Goues\\
       \affaddr{School of Computer Science}\\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{clegoues@cs.cmu.edu}
% 5th. author
\alignauthor
David Lo\\
\affaddr{Singapore Management University}\\
\affaddr{Singapore}\\
\email{davidlo@smu.edu.sg}       
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
%\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
%email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
%(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{February 2, 2016}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

\paragraph{How many files are changed to fix a bug?}
To answer this question, we identify all the bug fixing revisions in our dataset
and count the number of changed files. A file was changed if (1) it was a new
file; (2) or it was modified; (3) or it was deleted. In total, $52,052,571$
files got changed. Thus, on average each bug fixing revision changes 11.3 files.
This number is surprisingly high because most automatic program repair
techniques assume that bugs are local in most cases. There could be several
reasons behind this number. It could be the case that other software artifacts
(e.g.  documentation) are updated after bug fixing. It could also be the case
that bug fixing commit also contains changes that are not related to fixing bug
(e.g. refactoring, adding new features by creating new files). To better
understand why so many files are changed, we continue to ask:

\paragraph{What are the file types of those changed files?}
In Boa, each changed file is represented by a \emph{ChangedFile} node, and it
has a field called \emph{FileKind}, describing the type of each file. As
mentioned earlier, a \emph{ChangedFile} could be either a new file or an
existing file (if the bug fixing revision modified or deleted the file).
Table~\ref{tbl:fileType} shows the file types that Boa could identify. To answer
this question, we iterate all the changed files in each bug fixing revision and
count the total number for each file type. Results are shown in
Table~\ref{tbl:fileType}.

@@

\emph{SOURCE\_JAVA\_ERROR} file kind represents files that Boa failed to parse.
\emph{JLS2} refers to Java files that are written in all versions of Java
languages up to and including J2SE 1.4. \emph{JLS3} refers to J2SE 1.5, and JLS4
refers to J2SE 1.6 and 1.7. Boa documentation does not explicitly explain what
is \emph{TEXT} and what is \emph{UNKNOWN}. Our assumption is that \emph{TEXT}
refers to files whose file name ends with ``.txt'', and \emph{UNKNOWN} refers to
all other kinds of files (e.g. .c file, .cpp files).

From Table~\ref{tbl:fileType}, we can see that \emph{TEXT} files and
\emph{BINARY} files are changed least frequently. This is not surprising because
\emph{TEXT} files are usually documentation files, and \emph{BINARY} files
should only be changed in rare case (e.g.  milestone, new release, dependency
update). If we consider all 4 kinds of Java files as one kind, then on average
each bug fixing revision changes $4.47$ Java files. But still, it does not
conform to our assumption of bug locality. \emph{XML} files in Java projects
usually represents build files, thus changing one build file after each bug
fixing is not surprising.  What surprises us is that \emph{UNKNOWN} files are
changed most frequently. Note that all the projects in our dataset are Java
projects, so it is reasonable to assume that these files are not source files
that written in other programming languages.  One possible explanation is that
they are test resources. However, due to the limitations of Boa, we can not look
into the content of these \emph{UNKNOWN} files.

Since Boa does not distinguish Java files that related to functionality and Java
files that related to testing, it could be the case that among those $4.47$
changed Java files, some of them are related to testing, such as modifying the
test cases and creating new test classes. To better understand what changes are
usually made to Java source code files, we continue to ask:

\paragraph{
How often do developers introduce new classes, methods, fields, variables in
	bug fixing commits?}
This research question is interesting because most automatic program repair
approaches often do not create new classes/methods/fields/variables to fix bugs.
Most approaches guess and synthesize the patches based on existing source code,
and direct translations of approaching from C to Java do not naturally consider
object-oriented concerns, like fields or classes.

Table~\ref{tbl:new} shows our analysis results. This table shows that
introducing new classes and new variables are rare. This is good news for
automatic program repair approaches. On average, we still have $0.69$ new method
per bug fix. If we consider the possibility of creating new test method (i.e.
methods that annotated by @Test if JUnit is used), this number is not that
scary. However, what surprises us is that $1.32$ fields are created per bug fix.
This indicates that automatic program repair approaches should pay attention to
the state of the class when fixing Java programs.

\begin{table}
\centering
  \begin{tabular}{| c | r | r |}
  \hline
  Introducing & Total Number & Average \\ \hline \hline
  Class & 729,201 & 0.16 \\ \hline
  Methods & 3,186,867 & 0.69 \\ \hline
  Fields & 6,076,646 & 1.32 \\ \hline
  Variables & 924,259 & 0.2 \\ \hline
  \end{tabular}
  \caption{Introducing new classes, methods, fields, variables to fix bugs}
  \label{tbl:new}
\end{table}

@@

\subsection{Replacements}
Because direct diffs are difficult to identify, we heuristically approximate the
extent to which one statement type appears to be ``replaced'' by another across
this dataset.  To do this, for each modified file, we count the number of
appearances of each statement type in the file pre- and post-commit.  We then
compare the results to see how many of each statement type was removed, and how
many inserted, to roughly characterize the types of replacement that happen at a
per-file level.  For each statement type (as tagged by the Boa infrastructure
for Java), we process the results as follows: 
if number of occurrences of one statement type decreased post-fix and the number of
another type increased, we say that
the first statement type was replaced by the second statement type for that
file.

It is important to notice that this analysis doesn't consider replacing an
statement kind for a different version of the same statement kind, since it is
counting the amount of appearances of each statement kind. So, if for example,
the developer modifies the condition inside of an if statement, the amount of if
statements in the file will remain the same.

The numbers in the slots of Table~\ref{tab:likeliness}, represent the percentage rate at which the
elements of the leftmost column were replaced by the elements of the top row. In
the first column we have the list of statements that are being replaced (faulty
statements that were removed), and on the top row we have the list of statements
for which that statement was replaced for (the new statement put instead). For
example, if we take the intersection of row 5 (DoStatement row) and column 2
(AsserStatement column) of the data, we have a 0.81, which means that Do statements were replaced by Assert statements 0.81\% of the times.

\subsubsection{Most common replacements}
Our results show that the most common replacement is when developers remove a
Return statement and insert an If statement. This showed up in 30,489 files in
the code database. The second most common replacement is removing an If
statement and insert a Return statement instead, which had 28,536 appearances in
the code database. Third to that, is to remove a Return statement and insert a
Throw statement.

\subsubsection{Least common replacements}
On the opposite side of the data, we can see that the least common replacement
was an Assert statement replacing a Type Declaration statement, which has zero appearances. The second least common replacement was
removing a Do statement and insert a Type Declaration statement, with one
appearance in our search, and on third place, removing a Label and insert
a Type Declaration Statement had one appearance in our search.

\subsection{Deletions and insertions}

\begin{figure*}[!t] \centering
	\includegraphics[height=0.3\textheight, width=1\textwidth]{StmtType} \caption{For each kind
	of statement, how many bug fixing revisions add/delete statements of the
same kind?} \label{fig:StmtType} \end{figure*}
\todo{I'm not sure the percentages in Figure~\ref{fig:StmtType} are visible, maybe moving the numbers so they don't overlap with the bars?}
For each bug fixing revision $r$ and each statement kind $k$, we count how many
statements of kind $k$ are there in revision $r$ and $r-1$ ($r-1$ means the
revision before revision $r$), and then compare the two numbers.

From Figure~\ref{fig:StmtType} \todo{fix the small fonts in this graph}, we can
see that expression, if, return, for and try catch statements are added most
often. On the other hand, the same types of statements are also deleted most
often. This finding indicates that most bugs were fixed by changing the control
flow of the program using if, return, for, while, and try catch statements.

For each bug fixing revision, the results discussed above consider both newly
added files and existing files. Since most automatic program repair approaches
do not generate new files to fix bugs, the rest of this section only considers
files that were already in the repositories when developers fixed bugs. Our
analysis shows that there are 46,301,429 of them.

We found that the most common $Replacer$ statement kind (the statement
kind that most commonly replaces others) is the \textbf{If Statement} with
101,366 appearances; and the least common $Replacer$, is the \textbf{Type Declaration
  Statement} with 447 appearances.

The most common $Replacee$ (the most likely statement kind to be replaced
by another statement kind) is the \textbf{Return Statement} with 111,938
appearances; and the least common $Replacee$ is the \textbf{Type Declaration Statement}
with 399 appearances.

\subsection{Applicability to Automatic Error Repair}
This study is directly applicable to automatic software repair. Most common
tools for automatic software repair~\cite{kim2013,weimer2009,legoues2012}
first find a faulty statement using several different fault localization
techniques~\cite{fry2010}, and once they have that faulty statement, they apply
different kinds of edits to it, usually at random. Which makes it equally likely
to replace an statement for another without looking at the kinds of statements
being replaced.

With this study, we are able to make a more accurate guess, when having a known
faulty statement, of which statements more commonly replace that faulty one to build a successful patch based on the data from Table~\ref{tab:likeliness}. Similarly with Figure~\ref{fig:StmtType}, we can have an informed guess of how often programmers insert or delete that particular kind of statement.

@@

Some threats to our study are due to the limitation of Boa. For example, we rely
on Boa to identify bug fixing revisions, but it is unclear whether those
identified revisions are actually related to bug fixing. Chances are that there
exist some other bug fixing revisions, but Boa could not identify them. It is
also possible that bug fixing revisions identified by Boa actually include other
code changes, such as implementing new features, refactoring the code base.
After all, identifying bug fixing revisions from revision history is still an
open problem. Another big limitation of Boa is that there is no easy to get a
precise statement-level diff between two revisions. Thus, our template matching
and analysis of code changes (by counting each statement kind) could be
inaccurate. However, our results should provide upper bounds for the questions
we investigated.

>>>>

\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors section.
%
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
\alignauthor
Mauricio Soto\\
       \affaddr{School of Computer Science}\\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{mauriciosoto@cmu.edu}
       % 3rd. author
\alignauthor Ferdian Thung\\
       \affaddr{Singapore Management University}\\
       \affaddr{Singapore}\\
       \email{ferdiant.2013@smu.edu.sg}
% 2nd. author
\alignauthor Chu-Pan Wong   \\
       \affaddr{School of Computer Science}\\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{chupanw@cs.cmu.edu}       
\and  % use '\and' if you need 'another row' of author names
% 4th. author
\alignauthor
Claire Le Goues\\
       \affaddr{School of Computer Science}\\
       \affaddr{Carnegie Mellon University}\\
       \affaddr{Pittsburgh, PA}\\
       \email{clegoues@cs.cmu.edu}
% 5th. author
\alignauthor
David Lo\\
\affaddr{Singapore Management University}\\
\affaddr{Singapore}\\
\email{davidlo@smu.edu.sg}       
}
% There's nothing stopping you putting the seventh, eighth, etc.
% author on the opening page (as the 'third row') but we ask,
% for aesthetic reasons that you place these 'additional authors'
% in the \additional authors block, viz.
%\additionalauthors{Additional authors: John Smith (The Th{\o}rv{\"a}ld Group,
%email: {\texttt{jsmith@affiliation.org}}) and Julius P.~Kumquat
%(The Kumquat Consortium, email: {\texttt{jpkumquat@consortium.net}}).}
\date{February 2, 2016}
% Just remember to make sure that the TOTAL number of authors
% is the number that will appear on the first page PLUS the
% number that will appear in the \additionalauthors section.

@@

\subsubsection*{How many files are changed to fix a bug?} \vspace{-0.3em}

A file was changed if (1) it was a new file; (2) or it was modified; (3) or it
was deleted. In total, $52,052,571$ files got changed. Thus, on average each bug
fixing revision changes $11.3$ files.  This number is surprisingly high because
most automatic program repair techniques assume that bugs are local in most
cases. There could be several reasons behind this number. It could be the case
that other software artifacts (e.g.  documentation) are updated after bug
fixing. It could also be the case that bug fixing commit also contains changes
that are not related to fixing bug (e.g. refactoring, adding new features by
creating new files). To better understand why so many files are changed, we
continue to ask:

\subsubsection*{What are the file types of those changed files?} \vspace{-0.3em}

To answer this question, we iterate all the changed files in each bug fixing
revision and count the total number for each file type. Results are shown in
Table~\ref{tbl:fileType}.

@@

SOURCE\_JAVA\_ERROR file kind represents files that Boa failed to parse. JLS2,
JLS3, and JLS4 represent Java files written in different J2SE version. Boa
documentation does not explicitly explain what is TEXT and what is UNKNOWN. From
Table~\ref{tbl:fileType}, we can see that TEXT files and BINARY files are
changed least frequently. This is not surprising because TEXT files are usually
documentation files, and BINARY files should only be changed in rare case (e.g.
milestone, new release, dependency update). If we consider all 4 kinds of Java
files as one kind, then on average each bug fixing revision changes $4.47$ Java
files. But still, it does not conform to our assumption of bug locality. XML
files in Java projects usually represents build files, thus changing one build
file after each bug fixing is not surprising.  What surprises us is that UNKNOWN
files are changed most frequently. Note that all the projects in our dataset are
Java projects, so it is reasonable to assume that these files are not source
files that written in other programming languages.  One possible explanation is
that they are test resources. However, due to the limitations of Boa, we can not
look into the content of these files.

It could be the case that among those $4.47$ changed Java files, some of them
are related to testing, such as modifying the test cases and creating new test
classes. To better understand what changes are usually made to Java source code
files, we continue to ask:

\subsubsection*{How often do developers introduce new classes, methods, fields,
	local variables in
bug fixing commits?} \vspace{-0.3em}

This question is interesting because most automatic program repair approaches
often do not create new classes/methods/fields/local variables to fix bugs.
Most approaches guess and synthesize the patches based on existing source code,
and direct translations of approaches from C to Java do not naturally consider
object-oriented concerns, like fields or classes.

Table~\ref{tbl:new} shows our analysis results. It shows that introducing new
classes and new variables are rare, which is good news for automatic program
repair approaches. On average, we still have $0.69$ new method per bug fix. If
we consider the possibility of creating new test method (i.e.  methods that
annotated by @Test if JUnit is used), this number is not that scary. However,
what surprises us is that $1.32$ fields are created per bug fix.  This indicates
that automatic program repair approaches should pay attention to the state of
the class when fixing Java programs.

\begin{table}
\centering
  \begin{tabular}{| c | r | r |}
  \hline
  Introducing & Total Number & Average \\ \hline \hline
  Class & 729,201 & 0.16 \\ \hline
  Methods & 3,186,867 & 0.69 \\ \hline
  Fields & 6,076,646 & 1.32 \\ \hline
  Variables & 924,259 & 0.2 \\ \hline
  \end{tabular}
  \caption{Introducing new classes, methods, fields, local variables to fix bugs}
  \label{tbl:new}
\end{table}

@@

\subsection{Replacements}

Because direct diffs are difficult to identify, we heuristically approximate the
extent to which one statement type appears to be ``replaced'' by another across
this dataset.  To do this, for each modified file, we count the number of
appearances of each statement type in the file pre- and post-commit.  We then
compare the results to see how many of each statement type was removed, and how
many inserted, to roughly characterize the types of replacement that happen at a
per-file level.  For each statement type (as tagged by the Boa infrastructure
for Java), we process the results as follows: if number of occurrences of one
statement type decreased post-fix and the number of another type increased, we
say that the first statement type was replaced by the second statement type for
that file. It is important to notice that this analysis doesn't distinguish the
replacement of the same statement kind, since we are counting the amount of
appearances of each statement kind.

The numbers in the slots of Table~\ref{tab:likeliness} \todo{shall we add \% in
this table?}, represent the percentage rate at which the elements of the
leftmost column were replaced by the elements of the top row. In the first
column we have the list of statements that are being replaced (faulty statements
that were removed), and on the top row we have the list of statements for which
that statement was replaced for (the new statement put instead). For example, if
we take the intersection of row 5 (DoStatement row) and column 2 (AsserStatement
column) of the data, we have a 0.81, which means that Do statements were
replaced by Assert statements 0.81\% of the times.

Our results show that the most common replacement is replacing Return statement
with If statement, which showed up in 30,489 files. The second most common
replacement is replacing an If statement with a Return statement, which had
28,536 appearances. We found that the most common \emph{Replacer} statement kind
(the statement kind that most commonly replaces others) is the \textbf{If
Statement} with 101,366 appearances; and the least common \emph{Replacer} is the
\textbf{Type Declaration Statement} with 447 appearances.

On the opposite side of the data, the least common replacement was an Assert
statement replacing a Type Declaration statement, which had zero appearances.
The second least common replacement were replacing a Do statement with a Type
Declaration statement and replacing a Label with a Type Declaration, both of
which had one appearance respectively in our search. The most common $Replacee$
(the most likely statement kind to be replaced by another statement kind) is the
\textbf{Return Statement} with 111,938 appearances; and the least common
$Replacee$ is the \textbf{Type Declaration Statement} with 399 appearances.

\subsection{Deletions and Insertions}

\begin{figure*}[!t] \centering \includegraphics[height=0.3\textheight,
	width=0.8\textwidth]{StmtType} \caption{For each kind of statement, how many
	bug fixing revisions add/delete statements of the same kind?}
	\label{fig:StmtType} \end{figure*}

For each bug fixing revision $r$ and each statement kind $k$, we count how many
statements of kind $k$ are there in revision $r$ and $r-1$ ($r-1$ means the
revision before revision $r$), and then compare the two numbers.

From Figure~\ref{fig:StmtType}, we can see that, expression, if, return, for and
try catch statements are added most often. Meanwhile, the same types of
statements are also deleted most often. This finding indicates that most bugs
were fixed by changing the control flow of the program.

\subsection{Applicability to Automatic Error Repair}

This study is directly applicable to automatic software repair. Most common
tools for automatic software repair~\cite{kim2013,weimer2009,legoues2012} first
find a faulty statement using several different fault localization
techniques~\cite{fry2010}, and once they have that faulty statement, they apply
different kinds of edits to it, usually at random. Which makes it equally likely
to replace an statement for another without looking at the kinds of statements
being replaced. With this study, we are able to make a more accurate guess based
on the phenomenon we observed from Table~\ref{tab:likeliness} and
Figure~\ref{fig:StmtType}.

@@

Some threats to our study are due to the limitation of Boa. For example, we rely
on Boa to identify bug fixing revisions, but it is unclear whether those
identified revisions are actually related to bug fixing. After all, identifying
bug fixing revisions from revision history is still an open problem. Another big
limitation of Boa is that there is no easy way to get a precise statement-level
diff between two revisions. Thus, our template matching and analysis of code
changes (by counting each statement kind) could be inaccurate. However, our
results should provide upper bounds for the questions we investigated.
